---
title: "Custom Adaptive Stepped-Wedge Trial Design"
author: "Sam Watson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Adaptive Stepped-Wedge Trial Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
library(acrt)
library(ggplot2)
library(Matrix)

# Load precomputed results for expensive operations
load("sw_vignette_precomputed.rda")
```

## Introduction

The built-in design constructors in `adaptiveCRT` --- `parallel_crt()`, `crossover_crt()`, and `stepped_wedge_crt()` --- cover a range of standard trial layouts. However, many real trials require bespoke designs that don't fit neatly into these templates. This vignette walks through a complete example of building a **custom adaptive stepped-wedge cluster randomised trial** with a binomial outcome, demonstrating:

- How to write a custom `model_builder` function
- How to wrap it in a design constructor
- How to run the adaptive analysis across a grid of stage 1 designs
- How to use Pareto analysis to compare competing designs
- How to conduct interim analyses and sensitivity checks
- How to visualise the stepped-wedge trial layout

The trial we consider has $k = 11$ clusters observed across several time periods with a binary outcome. Clusters cross over from control to intervention in a staggered fashion. At an interim analysis after $t_1$ periods, we may adapt the number of additional periods ($t_2$), the rollout rate ($r$), and the cluster-period sample size ($m_2$).

## The stepped-wedge layout

In a standard stepped-wedge design, all clusters begin under control and sequentially cross to the intervention. Our design divides clusters into three groups:

1. **Early switchers** (clusters $1, \ldots, t_1 - 1$): these have already crossed over by the end of stage 1 and remain on intervention throughout stage 2.
2. **Middle clusters** (clusters $t_1, \ldots, k - t_1$): these are mid-crossover. In stage 2, the rate $r$ governs how quickly they switch. Setting $r = 0$ freezes them in place; $r = 1$ continues the standard wedge.
3. **Late clusters** (already on control through stage 1): they begin switching only in stage 2.

The adaptive element is that, after observing the stage 1 data, we choose $t_2$, $r$, and $m_2$ to balance power and cost.

We can visualise this layout with a helper function. First, a function to generate the design data frame:

```{r generate-sw-design}
generate_sw_design <- function(k, t1, t2, r) {
  cl_mid <- (k - t1 + 1) / 2

  # Stage 1: standard wedge
  df1 <- glmmrBase::nelder(as.formula(paste0("~ cl(", k, ") * t(", t1, ")")))
  df1$int <- as.integer(df1$t > df1$cl)

  # Stage 2: middle clusters continue crossing at rate r
  df2 <- glmmrBase::nelder(as.formula(paste0("~ cl(", (k - t1 + 1), ") * t(", t2, ")")))
  if (r < 1e-6) {
    df2$int <- as.integer(df2$cl <= cl_mid)
  } else {
    df2$int <- as.integer(df2$t > cl_mid + (df2$cl - cl_mid) / r)
  }
  df2$t <- df2$t + t1
  df2$cl <- df2$cl + (t1 - 1)

  # Stage 2: early switchers remain on intervention
  df3 <- glmmrBase::nelder(as.formula(paste0("~ cl(", t1 - 1, ") * t(", t2, ")")))
  df3$int <- 1L
  df3$t <- df3$t + t1

  rbind(df1, df2, df3)
}
```

And a plotting function using `ggplot2`:

```{r plot-sw-fn}
plot_sw_design <- function(df, t1 = NULL, title = NULL, show_legend = TRUE) {

  `%||%` <- function(x, y) if (is.null(x)) y else x

  df$cl  <- factor(df$cl, levels = sort(unique(df$cl), decreasing = TRUE))
  df$t   <- factor(df$t, levels = sort(unique(df$t)))
  df$int <- factor(df$int, levels = c(0, 1), labels = c("Control", "Intervention"))

  p <- ggplot(df, aes(x = t, y = cl, fill = int)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    scale_fill_manual(
      values = c("Control" = "#f0f0f0", "Intervention" = "#2a9d8f"),
      name = "Status"
    ) +
    labs(
      x = "Time period", y = "Cluster",
      title = title %||% "Stepped-Wedge Trial Design"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid   = element_blank(),
      axis.text    = element_text(size = 10),
      axis.title   = element_text(size = 11, face = "bold"),
      legend.position = if (show_legend) "right" else "none",
      plot.title   = element_text(size = 13, face = "bold", hjust = 0.5)
    )

  if (!is.null(t1)) {
    p <- p +
      geom_vline(
        xintercept = t1 + 0.5, colour = "#e76f51",
        linewidth = 1.2, linetype = "dashed"
      ) +
      annotate(
        "text", x = t1 + 0.5, y = length(unique(df$cl)) + 0.7,
        label = "Interim", colour = "#e76f51",
        fontface = "bold", size = 3.5, hjust = 0.5
      )
  }

  p
}
```

Here is one example layout with $k = 11$ clusters, $t_1 = 4$ stage 1 periods, $t_2 = 5$ additional periods, and a rollout rate of $r = 1$:

```{r plot-single-design, fig.height = 4}
df_example <- generate_sw_design(k = 11, t1 = 4, t2 = 5, r = 1)
plot_sw_design(df_example, t1 = 4, title = "SW Design: k=11, t1=4, t2=5, r=1.0")
```

## Writing the model builder

The model builder is the core function that computes the efficient information decomposition for a given combination of stage 1 and stage 2 parameters. It receives `design_params` (a single-row data frame of stage 2 parameters) and `fixed_params` (a list containing stage 1 parameters, variance components, and a model cache). It must return a list including `I1_eff`, `I2_eff`, `I_eff`, `w1`, `w2`, and `b1`.

Because this is a binomial outcome, we need helper functions to convert between the probability scale and the logit scale:

```{r helpers}
icc_var_par <- function(baseline, icc) {
  v1 <- baseline * (1 - baseline)
  v2 <- v1 * (icc / (1 - icc))
  tausq <- v2 / (baseline^2 / ((1 + exp(log(baseline / (1 - baseline))))^2))
  tausq
}

to_logit_scale <- function(baseline, eff_size) {
  b0 <- log(baseline / (1 - baseline))
  b1 <- log((baseline + eff_size) / (1 - baseline - eff_size)) - b0
  c(b0, b1)
}
```

Now the model builder itself. Note the use of caching: since this function is called many times during the optimisation, we only rebuild the `glmmrBase::Model` object when the cluster structure changes (i.e. when `t2` or `r` changes), and update weights when only $m_2$ changes.

```{r model-builder, eval = FALSE}
sw_model_builder <- function(design_params, fixed_params) {

  cache <- fixed_params$cache

  # --- Extract parameters ---
  k   <- fixed_params$k
  t1  <- fixed_params$t1
  t2  <- design_params$t2
  r   <- design_params$r
  m1  <- fixed_params$m1
  m2  <- design_params$m2
  ar  <- fixed_params$cac

  # --- Generate design data frame ---
  generate_sw <- function(k, t1, t2, r) {
    cl_mid <- (k - t1 + 1) / 2
    df1 <- glmmrBase::nelder(as.formula(paste0("~ cl(", k, ") * t(", t1, ")")))
    df1$int <- as.integer(df1$t > df1$cl)

    df2 <- glmmrBase::nelder(as.formula(paste0("~ cl(", (k - t1 + 1), ") * t(", t2, ")")))
    df3 <- glmmrBase::nelder(as.formula(paste0("~ cl(", t1 - 1, ") * t(", t2, ")")))
    df3$int <- 1L
    df3$t <- df3$t + t1

    if (r < 1e-6) {
      df2$int <- as.integer(df2$cl <= cl_mid)
    } else {
      df2$int <- as.integer(df2$t > cl_mid + (df2$cl - cl_mid) / r)
    }
    df2$t <- df2$t + t1
    df2$cl <- df2$cl + (t1 - 1)

    rbind(df1, df2, df3)
  }

  df <- generate_sw(k, t1, t2, r)
  df$n <- m1
  df[df$t > t1, "n"] <- m2

  # --- Convert to logit scale ---
  beta <- to_logit_scale(fixed_params$baseline, fixed_params$delta)
  v1   <- icc_var_par(fixed_params$baseline, fixed_params$icc)

  T_total <- t1 + t2

  # --- Build or update model (with caching) ---
  cache_key <- paste(k, t1, t2, r, sep = "_")

  if (is.null(cache$mod) || cache$cache_key != cache_key) {
    cache$mod <- glmmrBase::Model$new(
      ~ int + factor(t) + (1|gr(cl)*ar0(t)),
      data = df,
      family = binomial(),
      mean = c(beta, rep(0, T_total - 1)),
      covariance = c(v1, ar),
      trials = df$n
    )
    cache$cache_key <- cache_key
  } else {
    glmmrBase:::Model__set_trials(cache$mod$.__enclos_env__$private$ptr, df$n)
    cache$mod$update_parameters(cov.pars = c(v1, ar))
  }

  # --- Compute marginal covariance and efficient information ---
  n1 <- sum(df$t <= t1)
  n2 <- nrow(df) - n1

  X <- cache$mod$mean$X
  D <- cache$mod$covariance$D
  Z <- cache$mod$covariance$Z

  Z_sp <- Matrix(Z, sparse = TRUE)
  D_sp <- Matrix(D, sparse = TRUE)
  S <- Diagonal(x = 1 / cache$mod$w_matrix()) + Z_sp %*% tcrossprod(D_sp, Z_sp)
  V <- as.matrix(S)

  idx1 <- which(df$t <= t1)
  idx2 <- which(df$t > t1)

  eff <- efficient_score_decomposition(X, V, idx1, idx2, j = 2)

  list(
    I1_eff  = eff$I1_eff,
    I2_eff  = eff$I2_eff,
    I_eff   = eff$I_eff,
    w1      = sqrt(eff$I1_eff / eff$I_eff),
    w2      = sqrt(eff$I2_eff / eff$I_eff),
    b1      = beta[2],
    n1 = n1, n2 = n2,
    m1 = m1, m2 = m2,
    N1_total = k * t1 * m1,
    N2_total = k * t2 * m2,
    N_total  = k * t1 * m1 + k * t2 * m2,
    k = k, t1 = t1, t2 = t2, r = r,
    T_total = T_total
  )
}
```

## Creating the design constructor

We wrap everything into a constructor function `sw_crt()` that creates a `crt_design` object. This specifies the stage 1 and stage 2 parameter grids, the resource formulas for cost accounting, and the model builder.

```{r constructor, eval = FALSE}
sw_crt <- function(icc, cac = 0.8, delta, baseline,
                   k, m1, t1,
                   t2 = 1:10,
                   r = seq(0, 1.2, by = 0.2),
                   m2 = seq(20, 100, by = 20),
                   rho = 30) {

  spec <- crt_design_spec(
    stage1_params  = c("k", "t1", "m1"),
    stage2_params  = c("t2", "r", "m2"),
    resources = list(
      n_s1       = ~ k * t1 * m1,
      n_s2       = ~ k * t2 * m2,
      periods_s1 = ~ k * t1,
      periods_s2 = ~ k * t2
    ),
    cost_structure = list(
      weights = c(n = 1, periods = "rho"),
      stage2_resources = c("n_s2", "periods_s2")
    ),
    model_builder = sw_model_builder,
    n_arms = 1,
    design_type = "stepped_wedge"
  )

  structure(
    list(
      spec = spec,
      fixed_params = list(
        icc = icc, cac = cac, delta = delta,
        baseline = baseline, k = k
      ),
      stage1_grid    = expand.grid(t1 = t1, m1 = m1),
      stage2_grid_fn = function(s1) expand.grid(t2 = t2, r = r, m2 = m2),
      rho = rho
    ),
    class = c("sw_crt", "crt_design")
  )
}
```

## Exploring the design space

We first search over a grid of stage 1 configurations ($t_1 \in \{3,4,5,6\}$, $m_1 \in \{10,20,\ldots,60\}$) to understand how stage 1 choices affect the adaptive design's operating characteristics.

```{r design-grid, eval = FALSE}
design <- sw_crt(
  icc      = 0.06,
  cac      = 0.8,
  delta    = -0.07,
  baseline = 0.2,
  k        = 11,
  t1       = 3:6,
  m1       = seq(10, 60, by = 10),
  rho      = 15
)

results <- adaptive_analysis(
  design,
  target_power = 0.8,
  verbose = TRUE,
  tol = 0.01,
  method = "cost_cap"
)
```

The `summary()` method prints a table of results for each stage 1 configuration, including the calibrated $\lambda$, expected sample size, maximum sample size, expected cost, and decision probabilities.

```{r summary-grid, echo = FALSE}
cat(results_summary, sep = "\n")
```

### Plotting operating characteristics

Several plot types are available for comparing designs across the grid:

```{r plot-en, eval = FALSE}
plot(results, type = "EN")
```

```{r plot-en-show, echo = FALSE, fig.cap = "Expected total sample size across stage 1 configurations."}
print(p_EN)
```

```{r plot-nmax, eval = FALSE}
plot(results, type = "Nmax")
```

```{r plot-nmax-show, echo = FALSE, fig.cap = "Maximum sample size across stage 1 configurations."}
print(p_Nmax)
```

### Pareto analysis

In practice, we often face trade-offs: a design with low expected cost may have a high maximum cost, or vice versa. The `find_pareto()` function identifies the Pareto frontier across two or more objectives.

```{r pareto, eval = FALSE}
pareto <- find_pareto(
  results,
  objectives = list(E_cost = "min", max_cost = "min")
)

plot(results, type = "pareto",
     objectives = list(E_cost = "min", max_cost = "min"))
```

```{r pareto-show, echo = FALSE, fig.cap = "Pareto frontier of expected cost vs maximum cost."}
print(p_pareto)
```

## Single-design analysis

Once a promising stage 1 configuration has been identified, we can fix it and examine the adaptive operating characteristics in detail. Here we use $t_1 = 5$ and $m_1 = 30$:

```{r single-design, eval = FALSE}
design_single <- sw_crt(
  icc      = 0.06,
  cac      = 0.8,
  delta    = -0.07,
  baseline = 0.2,
  k        = 11,
  t1       = 5,
  m1       = 30,
  rho      = 15
)

results_single <- adaptive_analysis(
  design_single,
  target_power = 0.8,
  tol = 0.01,
  method = "cost_cap"
)
```

### Decision rules

The decision rule maps each possible value of the stage 1 test statistic $z_1$ to an action: stop for efficacy, stop for futility, or continue with a specific stage 2 design. Because the treatment effect here is negative (a reduction in the outcome rate), efficacy corresponds to large negative $z_1$.

```{r decision-rules, eval = FALSE}
plot(results_single, type = "decision")
```

```{r decision-show, echo = FALSE, fig.cap = "Decision regions as a function of the stage 1 test statistic."}
print(p_decision)
```

The decision rules can also be extracted as a data frame:

```{r rules-df}
head(rules[, c("z1", "continue", "cp", "m2", "t2", "r")])
```

## Interim analysis

At the interim, the observed $z_1$ and any updated nuisance parameter estimates are fed into `interim_analysis()` to obtain the recommended stage 2 design.

```{r interim, eval = FALSE}
interim_analysis(
  results_single,
  z1_obs    = -1.7,
  theta_hat = list(icc = 0.001)
)
```

```{r interim-show, echo = FALSE}
cat(interim_output, sep = "\n")
```

### Sensitivity analysis

We may not be confident in the nuisance parameter estimates. The `interim_sensitivity()` function evaluates the adaptive design under several scenarios, showing how the stage 2 recommendation changes:

```{r sensitivity, eval = FALSE}
sens <- interim_sensitivity(
  results_single,
  scenarios = list(
    "Planned (ICC = 0.06)"         = list(icc = 0.06),
    "Favourable (ICC = 0.02)"      = list(icc = 0.02),
    "Very favourable (ICC = 0.01)" = list(icc = 0.01)
  )
)

summary(sens)
```

```{r sens-summary-show, echo = FALSE}
cat(sens_summary, sep = "\n")
```

```{r sens-plots, eval = FALSE}
plot(sens, type = "decision")
plot(sens, type = "cp")
plot(sens, type = "all")
```

```{r sens-decision-show, echo = FALSE, fig.cap = "Decision rules under different ICC scenarios."}
print(p_sens_decision)
```

```{r sens-cp-show, echo = FALSE, fig.cap = "Conditional power curves under different ICC scenarios."}
print(p_sens_cp)
```

```{r sens-all-show, echo = FALSE, fig.cap = "Combined sensitivity analysis panel."}
print(p_sens_all)
```

## Comparing stage 2 adaptations visually

A useful way to communicate the adaptive design to a clinical audience is to show what the trial layout would look like under different interim outcomes. Suppose the three ICC scenarios from the sensitivity analysis lead to different stage 2 designs:

```{r compare-layouts, fig.height = 4, fig.width = 10}
plot_sw_designs_compare <- function(designs, t1 = NULL) {
  df_list <- lapply(names(designs), function(nm) {
    d <- designs[[nm]]
    d$design <- nm
    d
  })
  df_all <- do.call(rbind, df_list)

  df_all$cl  <- factor(df_all$cl, levels = sort(unique(df_all$cl), decreasing = TRUE))
  df_all$t   <- factor(df_all$t, levels = sort(unique(df_all$t)))
  df_all$int_label <- factor(df_all$int, levels = c(0, 1),
                             labels = c("Control", "Intervention"))
  df_all$design <- factor(df_all$design, levels = names(designs))

  p <- ggplot(df_all, aes(x = t, y = cl, fill = int_label)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    scale_fill_manual(
      values = c("Control" = "#e8e8e8", "Intervention" = "#2a9d8f"),
      name = "Status"
    ) +
    facet_wrap(~ design, ncol = length(designs)) +
    labs(x = "Time period", y = "Cluster") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold", size = 11),
      legend.position = "bottom"
    )

  if (!is.null(t1)) {
    p <- p + geom_vline(
      xintercept = t1 + 0.5, colour = "#e76f51",
      linewidth = 1, linetype = "dashed"
    )
  }

  p
}

designs <- list(
  "ICC = 0.06 (t2=7, r=1.05)" = generate_sw_design(11, 5, 7, r = 1.05),
  "ICC = 0.02 (t2=3, r=0.70)" = generate_sw_design(11, 5, 3, r = 0.7),
  "ICC = 0.01 (t2=1, r=0.00)" = generate_sw_design(11, 5, 1, r = 0.0)
)

plot_sw_designs_compare(designs, t1 = 5)
```

Under the planned ICC of 0.06, the trial extends by 7 additional periods with rapid rollout. If the ICC turns out to be much smaller than expected, the trial can stop after just one additional period --- a substantial saving in time and resources, while still maintaining the target power.

## Summary

This vignette has demonstrated the full workflow for building a custom adaptive stepped-wedge design:

1. **Define the trial layout** with a design-generating function.
2. **Write a model builder** that computes efficient information via `efficient_score_decomposition()`, using model caching for performance.
3. **Wrap it in a constructor** using `crt_design_spec()` to specify resources and cost structures.
4. **Explore the design space** across stage 1 configurations with `adaptive_analysis()` and `find_pareto()`.
5. **Examine a single design** in detail with decision rule plots and tables.
6. **Conduct interim analyses** with `interim_analysis()` and assess robustness with `interim_sensitivity()`.
7. **Communicate the design** to collaborators with layout visualisations showing how the trial adapts.

For further details on the custom design interface, run `custom_design_help()` in R, or see the package documentation.
